---
title: "STAT685_ODFlow_final"
author: "Huiting Sheng, James Cadena, Mike Pfahler, Thomas Nadobny"
date: "05/08/2021"
output: html_document
---

# Import libraries 
```{r, message=FALSE}
# load package, install package if it is not installed
packages = c("data.table","sf", "bit64", "rjson","reshape2", "od","tidyverse","stplanr","ggmap","plotly","fields",
             "BBmisc","bipartite", "bibloco","devtools","car", "gridExtra") 

package.check = lapply(packages,
                       FUN = function(x) {
                        if (!require(x, character.only = TRUE)) {
                          install.packages(x, dependencies = TRUE)
                          library(x, character.only = TRUE)
                        }
                       }
                      )
```

# Load data
```{r, message = F}
# set working directory
setwd("/Users/huitingsheng/Downloads/Github/Origin-Destination_Human_Mobility_Flow_Data_Analysis")

raw.df = fread('HT_sf_bg_2.csv') 
load("HT_district.Rdata") 

```


# Data PreProcessing

combine data and clean data by removing duplicate and NA records and converting destination_cbgs to standard json format
```{r, cache= T}
# join Ht_district with raws.df
df_combined = HT_district %>% select(GEOID) %>% left_join(raw.df %>% mutate(GEOID = as.character(GEOID)), by = 'GEOID')

# clean up the file. remove duplicate records and records have missing value, convert destination_cbgs into standard json format. add a new column for day of week. 
df = df_combined %>% na.omit %>% unique %>% mutate(DOW = lapply(as.Date(date), weekdays), destination_cbgs= str_replace_all(destination_cbgs,"\"\"\"\"\"\"\"", "")) 

```

Only extract Harris County and Fort Bend County data, and seperate into weekdays and weekends. For weekday data, we exclude holidays, which is New Year(1/1/2020) and Martin Luther King Jr. Day(01/20/2020). We also add row number to each record for later combination

```{r, cache = T}
weekdays = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
weekends = c("Saturday","Sunday")
counties = c("Harris County", "Fort Bend County")
`%notin%` = Negate(`%in%`)

# extract weekday data
data.weekdays = df %>% 
                filter(DOW %in% weekdays & County %in% counties & date %notin% c('2020-01-01','2020-01-20')) %>%
                rename(Origin = GEOID) %>% # rename GEOID to Origin
                mutate(id = row_number()) 

# extract weekend data
data.weekend = df %>% 
               filter(DOW %in% weekends & County %in% counties) %>%
               rename(Origin = GEOID) %>% # rename GEOID to Origin
               mutate(id = row_number()) 
```

PreProcessing data
```{r, cache = T}
# parsing json string into data table
json_parse_weekdays = lapply(data.weekdays$destination_cbgs, fromJSON) %>%melt
colnames(json_parse_weekdays) = c("Count", "Destination","id")

json_parse_weekend = lapply(data.weekend$destination_cbgs, fromJSON) %>%melt
colnames(json_parse_weekend) = c("Count", "Destination","id")

# join with HT_district to get the Destination geometry
json_parse_weekdays = json_parse_weekdays %>% inner_join(HT_district, by = c("Destination" = "GEOID")) %>% select(Count,Destination,id, geometry)
json_parse_weekend = json_parse_weekend %>% inner_join(HT_district, by = c("Destination" = "GEOID")) %>% select(Count,Destination,id, geometry)

# combine data with parsed json data and remove record that origin = destination
data.weekdays = data.weekdays %>% inner_join(json_parse_weekdays, by = 'id') %>% filter(Origin != Destination) %>% select(-c(destination_cbgs,id)) 
data.weekend = data.weekend %>% inner_join(json_parse_weekend, by = 'id') %>% filter(Origin != Destination) %>% select(-c(destination_cbgs,id)) 


# aggregation: add a new column "ratio", which indicate the ratio of destination device count to total leaving device count
data.weekdays.agg = data.weekdays %>% st_drop_geometry %>% 
                    select(Origin, Destination, Count, DeviceCount, completely_home_device_count) %>% 
                    group_by(Origin, Destination) %>% 
                    summarise(ratio = sum(Count)/sum(DeviceCount - completely_home_device_count))


data.weekend.agg = data.weekend %>% st_drop_geometry %>% 
                   select(Origin, Destination, Count, DeviceCount, completely_home_device_count) %>% 
                   group_by(Origin, Destination) %>% 
                   summarise(ratio = sum(Count)/sum(DeviceCount - completely_home_device_count))


# add distance between origin and destination
data.weekday.sf = data.weekdays.agg %>%
                  inner_join(HT_district, by = c("Origin" = "GEOID"), keep="geometry") %>%
                  inner_join(HT_district, by = c("Destination" = "GEOID")) %>% 
                  select(Origin, Destination, ratio, geometry.x, geometry.y ) 

data.weekday.distance = data.weekday.sf %>%
                        cbind(distance.mile=as.matrix(
                                          rdist.earth.vec(
                                                        st_coordinates(st_centroid(data.weekday.sf$geometry.x)),
                                                        st_coordinates(st_centroid(data.weekday.sf$geometry.y)), 
                                                        miles = T
                                                        ))) %>% 
                        select(Origin, Destination, ratio, distance.mile)


data.weekend.sf = data.weekend.agg %>%
                  inner_join(HT_district, by = c("Origin" = "GEOID"), keep="geometry") %>%
                  inner_join(HT_district, by = c("Destination" = "GEOID")) %>% 
                  select(Origin, Destination, ratio, geometry.x, geometry.y ) 

data.weekend.distance = data.weekend.sf %>%
                        cbind(distance.mile=as.matrix(
                                          rdist.earth.vec(
                                                        st_coordinates(st_centroid(data.weekend.sf$geometry.x)),
                                                        st_coordinates(st_centroid(data.weekend.sf$geometry.y)), 
                                                        miles = T
                                                        ))) %>% 
                        select(Origin, Destination, ratio, distance.mile)


```


generate OD Matrix for weekday and weekend
```{r, cache = T}
# generate OD Matrix for weekeday
OD_Matrix_weekday= data.weekdays.agg %>% od_to_odmatrix(attrib = 'ratio') %>% replace_na(0) %>% as.data.frame %>% 
                   select(sort(names(.))) %>% slice(order(rownames(.))) # sort the row name and column names by the GEOID

# generate OD Matrix for weekend
OD_Matrix_weekend= data.weekend.agg %>%  od_to_odmatrix(attrib = 'ratio') %>% replace_na(0) %>% as.data.frame %>% 
                   select(sort(names(.))) %>% slice(order(rownames(.))) # sort the row name and column names by the GEOID
  
```


set up a threshold, and generate a new matrix based on the threshold
```{r, cache = T}
threshold = 0.15
data.weekdays.threshold = data.weekdays.agg %>% filter(ratio >= threshold)
data.weekend.threshold = data.weekend.agg %>% filter(ratio >= threshold)

# generate a new OD matrix based on the threshold
Matrix_weekday_threshold = data.weekdays.threshold %>%  od_to_odmatrix(attrib = 'ratio') %>% replace_na(0) %>% as.data.frame %>% 
                           select(sort(names(.))) %>% slice(order(rownames(.))) # sort the row name and column names by the GEOID
Matrix_weekend_threshold = data.weekend.threshold %>%  od_to_odmatrix(attrib = 'ratio') %>% replace_na(0) %>% as.data.frame %>% 
                           select(sort(names(.))) %>% slice(order(rownames(.))) # sort the row name and column names by the GEOID
```



# Visulization

First, let take a look the most popular destination on weekdays and weekend
```{r}
# weekday: calculate the total destination device counts group by date and Destination
destination_weekday = data.weekdays %>% st_drop_geometry() %>% group_by(date, Destination) %>% summarise(total = sum(Count))

# weekday: calculate the daily destination device counts and add the geometry
destination_weekday_1 = HT_district %>% 
                        inner_join(destination_weekday %>% group_by(Destination) %>% summarise(total = sum(total)/21) %>% arrange(desc(total)), 
                                   by=c("GEOID" = "Destination"), keep="geometry") %>%
                        select(GEOID, total, geometry) %>% rename(Destination = GEOID)

# weekend: calculate the total destination device counts group by date and Destination
destination_weekend = data.weekend %>% st_drop_geometry() %>% group_by(date, Destination) %>% summarise(total = sum(Count))

# weekend: calculate the daily destination device counts and add the geometry
destination_weekend_1 = HT_district %>% 
                        inner_join(destination_weekend %>% group_by(Destination) %>% summarise(total = sum(total)/8) %>% arrange(desc(total)), 
                                   by=c("GEOID" = "Destination"), keep="geometry") %>%
                        select(GEOID, total, geometry) %>% rename(Destination = GEOID)

```


plot the most popular destination of weekday and weekend
```{r}
library(gridExtra)
# plot the most popular destination of weekday and weekend
bbox = c(left = -96.1, bottom = 29.3, right = -94.9, top = 30.2)
HSmap = get_stamenmap(bbox = bbox, zoom = 10)
destination_weekday_map = ggmap(HSmap)+
                          geom_sf(data = destination_weekday_1 %>% filter(total >= 1000), aes(fill = total), alpha=0.5, inherit.aes = FALSE) +
                          coord_sf(crs = st_crs(HT_district))  +
                          labs(title = "Most popular destination on weekday") +
                          theme(plot.title = element_text(size = 9, face="bold"),
                                legend.title = element_text(size = 7),
                                legend.text  = element_text(size = 5),
                                legend.key.size = unit(0.5, "lines"),
                                axis.text.x = element_text(angle = 90, size = 5),
                                axis.text.y = element_text(size = 7),
                                axis.title = element_text(size=7, face="bold")
                                )

destination_weekend_map = ggmap(HSmap) +
                          geom_sf(data = destination_weekend_1 %>% filter(total >= 1000), aes(fill = total), alpha=0.5, inherit.aes = FALSE) +
                          coord_sf(crs = st_crs(HT_district)) +
                          labs(title = "Most popular destination on Weekend") +
                          theme(plot.title = element_text(size = 9, face="bold"),
                                legend.title = element_text(size = 7),
                                legend.text  = element_text(size = 5),
                                legend.key.size = unit(0.5, "lines"),
                                axis.text.x = element_text(angle = 90, size = 5),
                                axis.text.y = element_text(size = 7),
                                axis.title = element_text(size=7, face= "bold")
                                )


grid.arrange(destination_weekday_map, destination_weekend_map, ncol=2)

```




## ratio vs distance 
```{r}
# weekday plot
ratio_distance_weekday = ggplot(data=data.weekday.distance, aes(x=distance.mile,y=ratio)) + geom_point() + xlim(0, 100) + ylim(0,0.5)+
                          labs(title = "Figure 1: Ratio vs Distance (mile) for weekday", x = "Distance(mile)", y = "ratio") +
                          theme(plot.title = element_text(size = 9, face="bold"),
                                legend.title = element_text(size = 7),
                                legend.text  = element_text(size = 5),
                                legend.key.size = unit(0.5, "lines"),
                                axis.text.x = element_text(size = 5),
                                axis.text.y = element_text(size = 7),
                                axis.title = element_text(size=7, face= "bold")
                                )

# weekend plot
ratio_distance_weekend = ggplot(data = data.weekend.distance, aes(x = distance.mile, y = ratio)) + geom_point() + xlim(0, 100) + ylim(0,0.5)+ 
                         labs(title = "Figure 2: Ratio vs Distance (mile) for weekend",  x = "Distance(mile)", y = "ratio") +
                         theme(plot.title = element_text(size = 9, face="bold"),
                                legend.title = element_text(size = 7),
                                legend.text  = element_text(size = 5),
                                legend.key.size = unit(0.5, "lines"),
                                axis.text.x = element_text( size = 5),
                                axis.text.y = element_text(size = 7),
                                axis.title = element_text(size=7, face= "bold")
                                )

grid.arrange(ratio_distance_weekday, ratio_distance_weekend, ncol=2, heights=c(2.5/4, 1.5/4))


# weekday + weekend plot
data.all = rbind(data.weekday.distance%>%mutate(cat = "weekday"),data.weekend.distance%>%mutate(cat = "weekend"))

ggplot(data=data.all, aes(x=distance.mile,y=ratio, color= cat)) +geom_point(alpha = 0.3)+ xlim(0, 100) + ylim(0,0.5) + labs(title = "Figure 3: Ratio vs Distance (mile) for weekday and weekend ", x = "Distance(mile)", y = "ratio")

distance_dist = ggplot(data= data.all, aes(x = distance.mile, fill = cat)) + geom_density(alpha = 0.3) +
                labs(title = "Figure 3: Distance Distribution",  x = "Distance (mile)") +
                theme(plot.title = element_text(size = 9, face="bold"),
                                  legend.title = element_text(size = 7),
                                  legend.text  = element_text(size = 5),
                                  legend.key.size = unit(0.5, "lines"),
                                  axis.text.x = element_text( size = 5),
                                  axis.text.y = element_text(size = 7),
                                  axis.title = element_text(size=7, face= "bold")
                                  )

            
ratio_dist = ggplot(data= data.all, aes(x = ratio, fill = cat)) + geom_density(alpha = 0.3) +
             labs(title = "Figure 4: Ratio Distribution",  x = "Ratio") +
             theme(plot.title = element_text(size = 9, face="bold"),
                                legend.title = element_text(size = 7),
                                legend.text  = element_text(size = 5),
                                legend.key.size = unit(0.5, "lines"),
                                axis.text.x = element_text( size = 5),
                                axis.text.y = element_text(size = 7),
                                axis.title = element_text(size=7, face= "bold")
                                )

grid.arrange(distance_dist, ratio_dist, ncol=2, heights=c(2.5/4, 1.5/4))
```



## Origin and Destination on Map for ratio greater than 0.1
```{r}

od.weekday.inner =  data.weekdays.threshold %>% filter(Origin %notin% c("480399900000","481679900000","480717106001","480719900000") & 
                                                       Destination %notin% c("480399900000","481679900000","480717106001","480719900000"))

od.weekend.inner =  data.weekend.threshold %>% filter(Origin %notin% c("480399900000","481679900000","480717106001","480719900000") & 
                                                       Destination %notin% c("480399900000","481679900000","480717106001","480719900000"))

                                       
l_weekday = od2line(flow = od.weekday.inner, zones = HT_district)
l_weekend = od2line(flow = od.weekend.inner, zones = HT_district)


library(leaflet)
leaflet() %>%
  addTiles() %>%
  addPolylines(data = l_weekday )

leaflet() %>%
  addTiles() %>%
  addPolylines(data = l_weekend )
```




# Model
## Gaussian for weekday
```{r}
#gaussian_matrix_weekday <- as.matrix( Matrix_weekday_threshold )
gaussian_weekday_adjusted = logit(data.weekdays.threshold$ratio, 
                                  percents=max(data.weekdays.threshold$ratio, na.rm = TRUE) > 1, 
                                  adjust=0)

data.weekdays.threshold$ratio = gaussian_weekday_adjusted

gaussian_matrix_weekday = data.weekdays.threshold %>% od_to_odmatrix(attrib = 'ratio') %>% replace_na(0) %>%
                          as.data.frame %>% select(sort(names(.))) %>% slice(order(rownames(.)))

params = list(alpha=2, kappa=1/var(as.numeric(gaussian_weekday_adjusted)), delta = 2, gamma=4, xi = mean(gaussian_weekday_adjusted))


iidx = NULL
jidx = NULL
x = NULL
for(i in 1:nrow(gaussian_matrix_weekday))
{
	for(j in 1:ncol(gaussian_matrix_weekday))
	{
		if(gaussian_matrix_weekday[i,j] != 0)
		{
			iidx = c(iidx,i)
			jidx = c(jidx,j)
			x = c(x,gaussian_matrix_weekday[i,j])
		}
	}
}
mat.dims = c(nrow(gaussian_matrix_weekday), ncol(gaussian_matrix_weekday))

fit_non_sparse_weekday = bibloco.fit(as.matrix(gaussian_matrix_weekday),
                                     model.type = "Gaussian", 
                                     model.params = params, 
                                     alg.type ="greedy",  
                                     init.type = 0, 
                                     n.runs = 10, 
                                     matrix.dims = mat.dims,
                                     init.groups = c(100, ncol(gaussian_matrix_weekday)), 
                                     n.restarts = 2,  
                                     greedy.fast = TRUE, 
                                     greedy.merge =TRUE, 
                                     keepICL = FALSE,
                                     row.index = iidx-1, 
                                     col.index = jidx-1,
                                     values = x,
                                     sparse = FALSE, 
                                     delta = 0.05,
                                     merge.thresh = 0., 
                                     n.margin = 2)

```

### weekday origination
```{r}

weekday_origin = rownames(gaussian_matrix_weekday[fit_non_sparse_weekday$plotting$rows$order,])
weekday_origin_group = c()
for (i in 1:fit_non_sparse_weekday$ngroups[1]){
  weekday_origin_group = c(weekday_origin_group, rep(i, length(fit_non_sparse_weekday$plotting$rows$cluster.members[[i]])))
}

origin_weekday = data.frame(weekday_origin, weekday_origin_group)

origin_weekday_sf = HT_district %>% inner_join(origin_weekday, by =c("GEOID" = "weekday_origin"))

```

plot weekday origination cluster
```{r}
bbox=c(left = -96.1, bottom = 29.2, right = -94.9, top = 30.2)  
HSmap <- get_stamenmap(bbox = bbox, zoom=10)
origin_weekday_map = ggmap(HSmap)+
                     geom_sf(data = origin_weekday_sf,  aes(fill = factor(weekday_origin_group)), alpha=0.5, inherit.aes = FALSE) +
                     coord_sf(crs = st_crs(HT_district)) + 
                     labs(fill = 'Clusters', title = 'Weekday Origin')

origin_weekday_map

```


### weekday destination
```{r}
weekday_dest = colnames(gaussian_matrix_weekday[,fit_non_sparse_weekday$plotting$columns$order])
weekday_dest_group = c()
for (i in 1:fit_non_sparse_weekday$ngroups[2]){
  weekday_dest_group = c(weekday_dest_group, rep(i,length(fit_non_sparse_weekday$plotting$columns$cluster.members[[i]])))
}

destination_weekday = data.frame(weekday_dest, weekday_dest_group)

destination_weekday_sf = HT_district %>% inner_join(destination_weekday, by = c("GEOID"="weekday_dest"))

```

plot weekday destination cluster
```{r}
bbox=c(left = -96.1, bottom = 29.2, right = -94.9, top = 30.2)  
HSmap <- get_stamenmap(bbox = bbox, zoom=10)
dest_weekday_map = ggmap(HSmap) +
                   geom_sf(data = destination_weekday_sf, aes(fill = factor(weekday_dest_group)), alpha=0.5, inherit.aes = FALSE) +
                   coord_sf(crs = st_crs(HT_district)) + 
                   labs(fill = 'Clusters', title='Weekday Destination')

dest_weekday_map

```


## Gaussian for weekend
```{r}
gaussian_weekend_adjusted = logit(data.weekend.threshold$ratio, 
                                  percents=max(data.weekend.threshold$ratio, na.rm = TRUE) > 1, adjust=0)

data.weekend.threshold$ratio = gaussian_weekend_adjusted

gaussian_matrix_weekend = data.weekend.threshold %>% od_to_odmatrix(attrib = 'ratio') %>% replace_na(0) %>%
                          as.data.frame %>% select(sort(names(.))) %>% 
                          slice(order(rownames(.)))

params = list(alpha = 2, kappa = 1/var(as.numeric(gaussian_weekend_adjusted)), delta = 2, gamma = 4, xi = mean(gaussian_weekend_adjusted))

iidx = NULL
jidx = NULL
x = NULL
for(i in 1:nrow(gaussian_matrix_weekend))
{
	for(j in 1:ncol(gaussian_matrix_weekend))
	{
		if(gaussian_matrix_weekend[i,j] != 0)
		{
			iidx = c(iidx,i)
			jidx = c(jidx,j)
			x = c(x,gaussian_matrix_weekend[i,j])
		}
	}
}
mat.dims = c(nrow(gaussian_matrix_weekend), ncol(gaussian_matrix_weekend))

fit_non_sparse_weekend = bibloco.fit(as.matrix(gaussian_matrix_weekend),
                                     model.type = "Gaussian", 
                                     model.params = params, 
                                     alg.type = "greedy",  
                                     init.type = 0, 
                                     n.runs = 10, 
                                     matrix.dims = mat.dims,
                                     init.groups = c(100,ncol(gaussian_matrix_weekend)), 
                                     n.restarts = 2,  
                                     greedy.fast = TRUE, 
                                     greedy.merge =TRUE, 
                                     keepICL = FALSE,
                                     row.index = iidx-1, 
                                     col.index = jidx-1,
                                     values = x,
                                     sparse = FALSE, 
                                     delta = 0.05,
                                     merge.thresh = 0., 
                                     n.margin = 2)

```

### weekend origination
```{r}

weekend_origin = rownames(gaussian_matrix_weekend[fit_non_sparse_weekend$plotting$rows$order,])
weekend_origin_group = c()
for (i in 1:fit_non_sparse_weekend$ngroups[1]){
  weekend_origin_group = c(weekend_origin_group, rep(i, length(fit_non_sparse_weekend$plotting$rows$cluster.members[[i]])))
}

origin_weekend = data.frame(weekend_origin, weekend_origin_group)

origin_weekend_sf = HT_district %>% inner_join(origin_weekend, by =c("GEOID" = "weekend_origin"))

```

```{r}

bbox=c(left = -96.1, bottom = 29.2, right = -94.9, top = 30.2)  
HSmap <- get_stamenmap(bbox = bbox,zoom = 10)
origin_weekend_map = ggmap(HSmap)+
                     geom_sf(data = origin_weekend_sf, aes(fill = factor(weekend_origin_group)), alpha = 0.5, inherit.aes = FALSE) +
                     coord_sf(crs = st_crs(HT_district)) + 
                     labs(fill = 'Clusters', title = 'Weekend Origin')

origin_weekend_map

```


### weekend destination
```{r}
weekend_dest = colnames(gaussian_matrix_weekend[,fit_non_sparse_weekend$plotting$columns$order])
weekend_dest_group = c()
for (i in 1:fit_non_sparse_weekend$ngroups[2]){
  weekend_dest_group = c(weekend_dest_group, rep(i,length(fit_non_sparse_weekend$plotting$columns$cluster.members[[i]])))
}

destination_weekend = data.frame(weekend_dest, weekend_dest_group)

destination_weekend_sf = HT_district %>% inner_join(destination_weekend, by = c("GEOID" = "weekend_dest"))

```

```{r}

bbox=c(left = -96.1, bottom = 29.2, right = -94.9, top = 30.2)  
HSmap = get_stamenmap(bbox = bbox,zoom = 10)
dest_weekend_map = ggmap(HSmap) +
                   geom_sf(data = destination_weekend_sf, aes(fill = factor(weekend_dest_group)), alpha=0.5, inherit.aes = FALSE) +
                   coord_sf(crs = st_crs(HT_district)) + 
                   labs(fill = 'Clusters', title = 'Weekend Destination')

dest_weekend_map

```